---
title: "Reproducible Research - PA-1"
author: "Ananth"
date: "Saturday, October 18, 2014"
output: html_document
---



```{r setoptions, echo=FALSE, results='hide'}

require(knitr)
opts_chunk$set(echo=TRUE, results='asis')

```

## Loading and preprocessing the data

Unzip and Read "activity.csv"" file from the default directory
```{r}
if(!file.exists('activity.csv')){
    unzip('activity.zip')
}
activity <- read.csv('activity.csv')
```

Add a new column "date.time"" by combining "date"" and "interval" columns.
```{r}
time <- formatC(activity$interval / 100, 2, format='f')
activity$date.time <- as.POSIXct(paste(activity$date, time),
                                 format='%Y-%m-%d %H.%M',
                                 tz='GMT')
```
Add a "time" column to analyze Means

```{r}
activity$time <- format(activity$date.time, format='%H:%M:%S')
activity$time <- as.POSIXct(activity$time, format='%H:%M:%S')
```

## What is mean total number of steps taken per day?

Calculate the mean number of steps for each day:
```{r}
total.steps <- tapply(activity$steps, activity$date, sum, na.rm=TRUE)
```

Draw a histogram for the total number of steps taken each day:

```{r histogram}
library(ggplot2)
qplot(total.steps, xlab='Total steps', ylab='Frequency')
```

Calculate mean and median for the total steps per day:
```{r}
mean(total.steps)
median(total.steps)
```



## What is the average daily activity pattern?

Calculate the mean steps for each five minute interval, and then put it in a data frame.
```{r}
mean.steps <- tapply(activity$steps, activity$time, mean, na.rm=TRUE)
daily.pattern <- data.frame(time=as.POSIXct(names(mean.steps)),
                            mean.steps=mean.steps)
```
Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r timeseriesplot}
library(scales)
ggplot(daily.pattern, aes(time, mean.steps)) + 
    geom_line() +
    xlab('Time of day') +
    ylab('Mean number of steps') +
    scale_x_datetime(labels=date_format(format='%H:%M'))
```

Which five minute interval has the highest mean number of steps?
```{r}
most <- which.max(daily.pattern$mean.steps)
format(daily.pattern[most,'time'], format='%H:%M')
```


## Imputing missing values
Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
summary(activity$steps)
```

Strategy is to use mean steps for a five-minute interval for the entire dataset.
```{r}
library(Hmisc)
activity.imputed <- activity
activity.imputed$steps <- with(activity.imputed, impute(steps, mean))
```
Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
total.steps.imputed <- tapply(activity.imputed$steps, 
                              activity.imputed$date, sum)
mean(total.steps)
mean(total.steps.imputed)
median(total.steps)
median(total.steps.imputed)
```

Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.

```{r histogram_imputed}
qplot(total.steps.imputed, xlab='Total steps', ylab='Frequency')
```

Imputing the missing data has increased the average number of steps. 

## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
day.type <- function(date) {
    if (weekdays(date) %in% c('Saturday', 'Sunday')) {
        return('weekend')
    } else {
        return('weekday')
    }
}

day.types <- sapply(activity.imputed$date.time, day.type)
activity.imputed$day.type <- as.factor(day.types)
```

Create a dataframe that holds the mean steps for weekdays and weekends.
```{r}
mean.steps <- tapply(activity.imputed$steps, 
                     interaction(activity.imputed$time,
                                 activity.imputed$day.type),
                     mean, na.rm=TRUE)
day.type.pattern <- data.frame(time=as.POSIXct(names(mean.steps)),
                               mean.steps=mean.steps,
                               day.type=as.factor(c(rep('weekday', 288),
                                                   rep('weekend', 288))))
```
Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)

```{r timeseries_daytype}
ggplot(day.type.pattern, aes(time, mean.steps)) + 
    geom_line() +
    xlab('Time of day') +
    ylab('Mean number of steps') +
    scale_x_datetime(labels=date_format(format='%H:%M')) +
    facet_grid(. ~ day.type)
```
